Comparative Analysis of Lion and AdamW Optimizers for Cross-Encoder Reranking with MiniLM, GTE, and ModernBERT

### 2025-04-25 cross_encoder/paper/main.tex
Refactor appendix to include detailed evaluation metrics and remove redundant sections
Updated some citations using xriv paper bibtex

\section{Introduction}
Information Retrieval (IR) systems aim to satisfy a user's information need, often expressed as a textual query, by returning a ranked list of relevant documents from a large collection. A common and effective architecture for modern search systems is the two-stage retrieve-and-rerank pipeline \cite{Lin2021PretrainedTF}. The first stage employs computationally efficient methods, such as BM25 \cite{robertson2009probabilistic} or dense vector retrieval \cite{karpukhin2020dense}, to quickly retrieve a candidate set of documents (typically hundreds or thousands) from the entire collection. The second stage then applies a more sophisticated and computationally expensive reranking model to reorder this smaller candidate set, aiming for higher precision at the top ranks.

Cross-encoder models, typically based on transformer architectures like BERT \cite{devlin2019bert}, have emerged as state-of-the-art rerankers \cite{nogueira2019passage, Nogueira2020Document}. Unlike bi-encoder models that encode queries and documents independently, cross-encoders process the query and a candidate document simultaneously (e.g., `[CLS] query [SEP] document [SEP]`). This allows for deep, token-level interaction modeling, leading to superior relevance estimation accuracy, albeit at a higher computational cost suitable only for the second stage. Some models like GTE \cite{li2023towards} and ModernBERT \cite{modernbert} support significantly longer input sequences (e.g., 8192 tokens), potentially benefiting the processing of longer passages or documents compared to models with shorter context windows like MiniLM \cite{wang2020minilm}.

Fine-tuning these large transformer models effectively is crucial. The choice of optimizer plays a key role in navigating the complex loss landscape and achieving optimal performance. Adam \cite{kingma2014adam} and its variant AdamW \cite{loshchilov2019decoupled} are widely used and generally effective optimizers for training deep neural networks. Recently, the Lion (Evo\textbf{L}ved S\textbf{i}gn M\textbf{o}me\textbf{n}tum) optimizer \cite{chen2023symbolic} was proposed, derived through symbolic mathematics and program search. It claims improved performance and memory efficiency compared to AdamW on various tasks, particularly image classification and vision-language models.

In this work, we investigate the applicability and effectiveness of the Lion optimizer for fine-tuning cross-encoder models specifically for the task of passage reranking in information retrieval. We compare its performance against the standard AdamW optimizer across three different base models with varying characteristics. Our contributions are:
\begin{itemize}
    \item We fine-tune three distinct transformer-based cross-encoder models (`microsoft/MiniLM-L12-H384-uncased`, `Alibaba-NLP/gte-multilingual-base`, and `answerdotai/ModernBERT-base`) on the large-scale MS MARCO passage dataset \cite{bajaj2016ms} using both Lion and AdamW optimizers.
    \item We utilize different training configurations optimized for specific models, including a lower learning rate (2e-6) and a Cosine Annealing scheduler for ModernBERT, contrasting with a higher rate (2e-5) and no scheduler for MiniLM and GTE.
    \item We evaluate the performance of the fine-tuned models across three training epochs on the TREC 2019 Deep Learning (DL) Track \cite{craswell2020overview} benchmark and the MS MARCO development set \cite{DBLP:journals/corr/NguyenRSGTMD16}.
    \item We provide a comparative analysis of the optimizers' impact on reranking effectiveness using standard IR metrics (NDCG@10, MAP, MRR@10, Recall@10, R-Prec, P@10).
    \item We utilize the Modal cloud platform \cite{modal_labs} for efficient GPU resource management (NVIDIA A100-80GB) and reproducible experimentation.
\end{itemize}
The rest of the paper is organized as follows: Section \ref{sec:related_work} discusses related work. Section \ref{sec:methodology} details the models, optimizers, and training approach. Section \ref{sec:experimental_setup} describes the experimental setup, datasets, and evaluation protocol. Section \ref{sec:results} presents and discusses the results. Finally, Section \ref{sec:conclusion} concludes the paper and suggests future work.

@@

\subsection{Cross-Encoder Reranking}
The use of BERT-based models for document reranking was popularized by Nogueira et al. \cite{nogueira2019passage, Nogueira2020Document}. They demonstrated that fine-tuning BERT as a cross-encoder on relevance labels (like those from MS MARCO) significantly outperforms traditional IR models and even bi-encoder approaches for reranking tasks. Subsequent work explored various transformer architectures \cite{Lin2021PretrainedTF}, training strategies \cite{gao2021complementing}, and efficiency improvements \cite{hofstatter2020improving}. Models like MiniLM \cite{wang2020minilm} offer a balance between effectiveness and efficiency through knowledge distillation. Multilingual models like GTE \cite{li2023towards} (General Text Embeddings) provide strong performance across various text tasks, including retrieval, and can be adapted for cross-encoding, supporting longer context lengths. Recently, models like ModernBERT \cite{modernbert} have incorporated architectural improvements such as Rotary Positional Embeddings (RoPE) \cite{su2023roformerenhancedtransformerrotary} and Flash Attention \cite{dao2022flashattentionfastmemoryefficientexact} to handle even longer sequences (up to 8192 tokens) efficiently and effectively. The `sentence-transformers` library \cite{reimers2019sentence} provides a convenient framework for training and using both bi-encoders and cross-encoders.

\subsection{Optimizers for Deep Learning}
Stochastic Gradient Descent (SGD) with momentum remains a fundamental optimizer, but adaptive learning rate methods like AdaGrad \cite{duchi2011adaptive}, RMSprop \cite{tieleman2012lecture}, and Adam \cite{kingma2014adam} often lead to faster convergence in practice for deep learning models. Adam combines momentum with adaptive scaling of learning rates based on estimates of first and second moments of the gradients. AdamW \cite{loshchilov2019decoupled} improves upon Adam by decoupling the weight decay regularization from the adaptive learning rate mechanism, often leading to better generalization. The Lion optimizer \cite{chen2023symbolic} takes a different approach, using only momentum tracking and a sign operation on the update, resulting in a simpler update rule (`update = sign(momentum) * lr`) and potentially requiring less memory due to not storing second moment estimates. Its effectiveness relative to AdamW has been shown primarily in vision and vision-language tasks, motivating its evaluation in the NLP/IR domain.

\subsection{Evaluation Benchmarks}
The MS MARCO (Microsoft MAchine Reading COmprehension) dataset \cite{bajaj2016ms, DBLP:journals/corr/NguyenRSGTMD16} has become a standard benchmark for training and evaluating deep learning models for passage retrieval and reranking. It contains real user queries from Bing and human-judged relevant passages. The TREC Deep Learning (DL) Tracks \cite{craswell2020overview, craswell2021overview} provide challenging test collections built upon MS MARCO, using queries with sparse relevance judgments derived from pooling top results from various participating systems. We use the TREC DL 2019 passage ranking dataset, a standard benchmark for evaluating reranking effectiveness. Evaluation is typically performed using tools like `trec\_eval` \cite{trec_eval_github}, which calculates various standard IR metrics. We specifically use the MS MARCO passage dataset's development split for evaluating MRR@10, providing a complementary perspective on a larger query set.

@@

\subsection{Training}
We fine-tune the cross-encoders using the MS MARCO passage ranking triplets dataset \cite{bajaj2016ms}, as processed by `sentence-transformers` \cite{reimers2019sentence}. The original dataset contains tuples of (query, positive passage, negative passage). We convert this into pairs `(query, positive\_passage)` with label 1 and `(query, negative\_passage)` with label 0. We train on approximately 2 million such pairs.
The training objective is to minimize the Binary Cross-Entropy (BCE) loss between the predicted relevance score $\hat{y}$ (output of the sigmoid function) and the true label $y$ (0 or 1). The BCE loss is defined as:
\[ \mathcal{L}_{BCE} = - [y \log(\hat{y}) + (1 - y) \log(1 - \hat{y})] \]

@@

\subsection{Datasets}
\begin{itemize}
    \item \textbf{Training:} We use the MS MARCO passage ranking triplets dataset \cite{bajaj2016ms}, processed into approximately 2 million query-passage pairs with binary labels using the `sentence-transformers` methodology.
    \item \textbf{Evaluation (Main):} We evaluate on the TREC 2019 Deep Learning Track passage ranking task \cite{craswell2020overview}, which contains 43 queries with graded relevance judgments (qrels).
    \item \textbf{Evaluation (MRR@10):} We evaluate Mean Reciprocal Rank at cutoff 10 (MRR@10) on the MS MARCO passage dataset's development split (available from Hugging Face Datasets \cite{DBLP:journals/corr/NguyenRSGTMD16}), containing a larger set of queries with binary relevance.
\end{itemize}

@@

\newpage
\appendix
\section*{Appendices}

@@

@@

\section{Metric Definitions}
\label{app:metric_defs}
Table~\ref{tab:metric_definitions} provides formal definitions of all evaluation metrics used in our experimental analysis.

>>>>

\section{Introduction}
Information Retrieval (IR) systems aim to satisfy a user's information need, often expressed as a textual query, by returning a ranked list of relevant documents from a large collection. A common and effective architecture for modern search systems is the two-stage retrieve-and-rerank pipeline \cite{hu2019retrievereadrerankendtoend}. The first stage employs computationally efficient methods, such as BM25 \cite{robertson2009probabilistic} or dense vector retrieval \cite{karpukhin2020dense}, to quickly retrieve a candidate set of documents (typically hundreds or thousands) from the entire collection. The second stage then applies a more sophisticated and computationally expensive reranking model to reorder this smaller candidate set, aiming for higher precision at the top ranks.

Cross-encoder models, typically based on transformer architectures like BERT \cite{devlin2019bert}, have emerged as state-of-the-art rerankers \cite{nogueira2020passagererankingbert, Nogueira2020Document}. Unlike bi-encoder models that encode queries and documents independently, cross-encoders process the query and a candidate document simultaneously (e.g., `[CLS] query [SEP] document [SEP]`). This allows for deep, token-level interaction modeling, leading to superior relevance estimation accuracy, albeit at a higher computational cost suitable only for the second stage. Some models like GTE \cite{li2023towards} and ModernBERT \cite{modernbert} support significantly longer input sequences (e.g., 8192 tokens), potentially benefiting the processing of longer passages or documents compared to models with shorter context windows like MiniLM \cite{wang2020minilm}.

Fine-tuning these large transformer models effectively is crucial. The choice of optimizer plays a key role in navigating the complex loss landscape and achieving optimal performance. Adam \cite{kingma2017adam} and its variant AdamW \cite{loshchilov2019decoupled} are widely used and generally effective optimizers for training deep neural networks. Recently, the Lion (Evo\textbf{L}ved S\textbf{i}gn M\textbf{o}me\textbf{n}tum) optimizer \cite{chen2023symbolic} was proposed, derived through symbolic mathematics and program search. It claims improved performance and memory efficiency compared to AdamW on various tasks, particularly image classification and vision-language models.

In this work, we investigate the applicability and effectiveness of the Lion optimizer for fine-tuning cross-encoder models specifically for the task of passage reranking in information retrieval. We compare its performance against the standard AdamW optimizer across three different base models with varying characteristics. Our contributions are:
\begin{itemize}
    \item We fine-tune three distinct transformer-based cross-encoder models (`microsoft/MiniLM-L12-H384-uncased`, `Alibaba-NLP/gte-multilingual-base`, and `answerdotai/ModernBERT-base`) on the large-scale MS MARCO passage dataset \cite{DBLP:journals/corr/NguyenRSGTMD16} using both Lion and AdamW optimizers.
    \item We utilize different training configurations optimized for specific models, including a lower learning rate (2e-6) and a Cosine Annealing scheduler for ModernBERT, contrasting with a higher rate (2e-5) and no scheduler for MiniLM and GTE.
    \item We evaluate the performance of the fine-tuned models across three training epochs on the TREC 2019 Deep Learning (DL) Track \cite{craswell2020overview} benchmark and the MS MARCO development set \cite{DBLP:journals/corr/NguyenRSGTMD16}.
    \item We provide a comparative analysis of the optimizers' impact on reranking effectiveness using standard IR metrics (NDCG@10, MAP, MRR@10, Recall@10, R-Prec, P@10).
    \item We utilize the Modal cloud platform \cite{modal_labs} for efficient GPU resource management (NVIDIA A100-80GB) and reproducible experimentation.
\end{itemize}
The rest of the paper is organized as follows: Section \ref{sec:related_work} discusses related work. Section \ref{sec:methodology} details the models, optimizers, and training approach. Section \ref{sec:experimental_setup} describes the experimental setup, datasets, and evaluation protocol. Section \ref{sec:results} presents and discusses the results. Finally, Section \ref{sec:conclusion} concludes the paper and suggests future work.

@@

\subsection{Cross-Encoder Reranking}
The use of BERT-based models for document reranking was popularized by Nogueira et al. \cite{nogueira2020passagererankingbert, Nogueira2020Document}. They demonstrated that fine-tuning BERT as a cross-encoder on relevance labels (like those from MS MARCO) significantly outperforms traditional IR models and even bi-encoder approaches for reranking tasks. Subsequent work explored various transformer architectures \cite{Lin2021PretrainedTF}, training strategies \cite{gao2021complementing}, and efficiency improvements \cite{hofstatter2020improving}. Models like MiniLM \cite{wang2020minilm} offer a balance between effectiveness and efficiency through knowledge distillation. Multilingual models like GTE \cite{li2023towards} (General Text Embeddings) provide strong performance across various text tasks, including retrieval, and can be adapted for cross-encoding, supporting longer context lengths. Recently, models like ModernBERT \cite{modernbert} have incorporated architectural improvements such as Rotary Positional Embeddings (RoPE) \cite{su2023roformerenhancedtransformerrotary} and Flash Attention \cite{dao2022flashattentionfastmemoryefficientexact} to handle even longer sequences (up to 8192 tokens) efficiently and effectively. The `sentence-transformers` library \cite{reimers2019sentence} provides a convenient framework for training and using both bi-encoders and cross-encoders.

\subsection{Optimizers for Deep Learning}
Stochastic Gradient Descent (SGD) with momentum remains a fundamental optimizer, but adaptive learning rate methods like AdaGrad \cite{duchi2011adaptive}, RMSprop \cite{tieleman2012lecture}, and Adam \cite{kingma2017adam} often lead to faster convergence in practice for deep learning models. Adam combines momentum with adaptive scaling of learning rates based on estimates of first and second moments of the gradients. AdamW \cite{loshchilov2019decoupled} improves upon Adam by decoupling the weight decay regularization from the adaptive learning rate mechanism, often leading to better generalization. The Lion optimizer \cite{chen2023symbolic} takes a different approach, using only momentum tracking and a sign operation on the update, resulting in a simpler update rule (`update = sign(momentum) * lr`) and potentially requiring less memory due to not storing second moment estimates. Its effectiveness relative to AdamW has been shown primarily in vision and vision-language tasks, motivating its evaluation in the NLP/IR domain.

\subsection{Evaluation Benchmarks}
The MS MARCO (Microsoft MAchine Reading COmprehension) dataset \cite{DBLP:journals/corr/NguyenRSGTMD16, DBLP:journals/corr/NguyenRSGTMD16} has become a standard benchmark for training and evaluating deep learning models for passage retrieval and reranking. It contains real user queries from Bing and human-judged relevant passages. The TREC Deep Learning (DL) Tracks \cite{craswell2020overview, craswell2021overview} provide challenging test collections built upon MS MARCO, using queries with sparse relevance judgments derived from pooling top results from various participating systems. We use the TREC DL 2019 passage ranking dataset, a standard benchmark for evaluating reranking effectiveness. Evaluation is typically performed using tools like `trec\_eval` \cite{trec_eval_github}, which calculates various standard IR metrics. We specifically use the MS MARCO passage dataset's development split for evaluating MRR@10, providing a complementary perspective on a larger query set.

@@

\subsection{Training}
We fine-tune the cross-encoders using the MS MARCO passage ranking triplets dataset \cite{DBLP:journals/corr/NguyenRSGTMD16}, as processed by `sentence-transformers` \cite{reimers2019sentence}. The original dataset contains tuples of (query, positive passage, negative passage). We convert this into pairs `(query, positive\_passage)` with label 1 and `(query, negative\_passage)` with label 0. We train on approximately 2 million such pairs.
The training objective is to minimize the Binary Cross-Entropy (BCE) loss between the predicted relevance score $\hat{y}$ (output of the sigmoid function) and the true label $y$ (0 or 1). The BCE loss is defined as:
\[ \mathcal{L}_{BCE} = - [y \log(\hat{y}) + (1 - y) \log(1 - \hat{y})] \]

@@

\subsection{Datasets}
\begin{itemize}
    \item \textbf{Training:} We use the MS MARCO passage ranking triplets dataset \cite{DBLP:journals/corr/NguyenRSGTMD16}, processed into approximately 2 million query-passage pairs with binary labels using the `sentence-transformers` methodology.
    \item \textbf{Evaluation (Main):} We evaluate on the TREC 2019 Deep Learning Track passage ranking task \cite{craswell2020overview}, which contains 43 queries with graded relevance judgments (qrels).
    \item \textbf{Evaluation (MRR@10):} We evaluate Mean Reciprocal Rank at cutoff 10 (MRR@10) on the MS MARCO passage dataset's development split (available from Hugging Face Datasets \cite{DBLP:journals/corr/NguyenRSGTMD16}), containing a larger set of queries with binary relevance.
\end{itemize}

@@

\appendix

@@

\subsection{Metric Definitions}
\label{app:metric_defs}
Table~\ref{tab:metric_definitions} provides formal definitions of all evaluation metrics used in our experimental analysis.

@@
